Ride at Your Own Risk: Insights into Toronto’s Bicycle Theft Hotspots

### 2024-11-12 paper/paper.Rmd
paper formmating

required_packages <- c("lubridate", "sf", "tidyverse", "readr", "here", "osmdata", "ggplot2")
for (p in required_packages) {
  if (!require(p, character.only = TRUE)) {
    install.packages(p, character.only = TRUE)
  }
}

library(lubridate)
library(tidyverse)
library(ggplot2)
library(osmdata)
library(readr)
library(here)
library(sf)

# you are welcome to use either the CSV file or the GeoJSON file
# GeoJSON format maybe useful if you would like to combine
# an external geospatial dataset
bike_thefts_csv <- read_csv(here("data/clean_data.csv")) |>
  mutate(
    occurence_date = as_date(
      OCC_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    report_date = as_date(
      REPORT_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    )
  )
bike_thefts_sf <- read_sf(here("data/Bicycle_Thefts_Open_Data.geojson")) |>
  mutate(
    occurence_date = as_date(
      OCC_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    report_date = as_date(
      REPORT_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    )
  )
```

@@

library(plotly)
library(readr)
library(dplyr)
library(here)

@@

# Initialize plot with a line trace for the entire data
fig <- plot_ly(data, x = ~OCC_DOW, y = ~Number_of_Thefts, type = 'scatter', mode = 'lines+markers',
               line = list(color = 'brown'), marker = list(color = 'blue', size = 8))

# Add points for each day with initial visibility settings
for (i in 1:nrow(data)) {
  fig <- fig %>%
    add_trace(
      x = data$OCC_DOW[i],
      y = data$Number_of_Thefts[i],
      type = 'scatter',
      mode = 'markers+text',
      text = paste("Thefts:", data$Number_of_Thefts[i]),
      textposition = "top center",
      marker = list(color = "yellow", size = 10),
      showlegend = FALSE,
      visible = ifelse(i == 1, TRUE, FALSE)
    )
}

@@

<div class="main-body">
  <p> Weekly bicycle theft patterns in Toronto is illustrated in @interactive_weekly_plot, revealing that thefts are not evenly distributed throughout the week. The data shows a peak on Fridays, with over <strong style="color: #FFC107;">4,900</strong> reported thefts, followed by a decline on Saturday and Sunday. This trend may reflect how people use their bikes during the week versus the weekend. Fridays might represent a combination of commuter bike usage and weekend planning, where people may leave their bikes in high-traffic or less secure areas, making them more vulnerable to theft.</p>

@@

<center>
```{r mothly_plot, include = TRUE, warning = FALSE, message = FALSE, fig.width=9.5, fig.height=5}
# Load necessary libraries
library(tidyverse)
library(ggridges)
library(cowplot)
library(here)

@@

@@

<center>
```{r animated_bar_plot, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=4.5, fig.height=2.3}
# Libraries
library(ggplot2)
library(gganimate)
library(readr)
library(here)
library(dplyr)

@@

<center>
```{r animated_line_plot, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=5, fig.height=3}
# Load necessary libraries
library(ggplot2)
library(gganimate)
library(dplyr)
library(hrbrthemes)
library(readr)
library(here)

@@

The year 2020, however, deviates from this regular trend due to the impact of the COVID-19 pandemic. As public transit options were considered risky, many Torontonians turned to bicycles as a safer and more flexible transportation option, resulting in a rise in bike ownership and driving up theft rates (@citeshapiro2021). Additionally, the pandemic stretched police resources, potentially limiting routine patrols and recovery efforts. The 2020 increase highlights how external factors, such as a public health crisis, can exacerbate bike theft trends and points to the importance of adaptable theft prevention strategies in response to changing urban transportation needs.

@@

# Load necessary libraries
library(dplyr)         # Provides functions for data manipulation.
library(leaflet)       # Enables interactive maps.
library(purrr)         # Contains functional programming tools, used here to walk through data.
library(readr)         # Provides functions to read rectangular data like CSV files.
library(here)          # Helps manage file paths in a more robust way.

# Load your dataset (adjust the file path accordingly)
map_data <- read_csv(here("data", "map_by_neighbourhoods_data.csv"), show_col_types = FALSE)
# Loads the CSV file from the specified path using the 'here()' function for consistent path management.
# 'show_col_types = FALSE' suppresses messages about the data types of each column.

# Modify the data frame, categorizing case_num column
map_data <- map_data %>% 
  mutate(
    mag.level = cut(case_num, c(100, 500, 1000, 1500),
                    labels = c('> 100 & <=500', '>500 & <=1000', '>1000 & <=1500'))
    # Creates a 'mag.level' column that categorizes 'case_num' into three levels based on specified ranges.
  ) %>%
  split(.$mag.level)  # Splits the dataset into a list of data frames by 'mag.level'.

# Initialize the leaflet map
l <- leaflet() %>%
  addProviderTiles(providers$Esri.OceanBasemap)
# Initializes a Leaflet map object and adds a base tile layer from the Esri Ocean Basemap provider.

# Add the markers to the map for each case_num level
names(map_data) %>%
  walk(function(df) {
    l <<- l %>% 
      addMarkers(
        data = map_data[[df]], lng = ~longitude, lat = ~latitude,
        label = ~paste0("Hood: ", hood_name, " | Stolen Place: ", stolen_place, " | Cost: ", cost, " | Status: ", status),
        popup = ~paste0(
          "Hood: ", hood_name, "<br>",
          "Stolen Place: ", stolen_place, "<br>",
          "Cost: ", cost, "<br>",
          "Status: ", status, "<br>",
          "Case in area: ", case_num
        ),
        group = df,
        clusterOptions = markerClusterOptions()
      )
  })
# Iterates over the names of the 'map_data' list (which are the different 'mag.level' categories).
# For each category, it adds markers to the map:
# - `data = map_data[[df]]`: Uses the split data for the current case_num level.
# - `lng = ~longitude, lat = ~latitude`: Sets longitude and latitude for marker placement.
# - `label`: Displays the 'case_num' value when markers are hovered over.
# - `popup`: Displays additional information (hood_num, stolen_place, cost, status, and case_num) when a marker is clicked.
# - `group = df`: Groups the markers by 'mag.level'.
# - `clusterOptions = markerClusterOptions()`: Enables clustering of markers for better visualization.

# Add layer controls and a minimap
l <- l %>%
  addLayersControl(
    overlayGroups = names(map_data),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addMiniMap(tiles = providers$Esri.OceanBasemap, width = 120, height = 80)
# Adds layer controls to allow switching between different 'mag.level' categories on the map.
# Adds a minimap showing a small overview of the area using the Esri Ocean Basemap.

# Display the Map
l
```
</center>

@@

<center>
```{r map_of_stolen_bike, echo=FALSE, warning=FALSE, message=FALSE}
# Load necessary libraries
library(ggplot2)
library(sf)
library(osmdata)
library(here)

@@

# Plot the map with bike theft locations data
bike_theft_plot <- ggplot() +
  geom_sf(data = streets, color = "gray", size = 0.3, alpha = 0.6) + # OSM street map layer
  geom_point(
    data = coordinates_data, aes(x = longitude, y = latitude),
    size = 0.5, color = "red"
  ) + # Bike theft points
  labs(
    title = "Map of Stolen Bike Cases in Toronto",
    x = "Longitude", y = "Latitude"
  ) +
  coord_sf(
    xlim = c(-79.6303856025883, -79.1180336515019),
    ylim = c(43.5822069220627, 43.8554571861712), expand = FALSE
  ) +
  theme_minimal() +
  theme(
    panel.background = element_blank(), # Remove panel background
    panel.grid.major = element_blank(), # Remove major grid lines
    panel.grid.minor = element_blank(), # Remove minor grid lines
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

bike_theft_plot
```
</center>

@@

# Load necessary libraries
library(dplyr)
library(readr)
library(ggplot2)
library(plotly)
library(sf)
library(crosstalk)
library(here)
library(osmdata)

@@

# Create ggplot object with the dataset and Toronto map as a background
ggplot_map <- ggplot() +
  geom_sf(data = streets, color = "lightgray", size = 0.3, alpha = 0.6) +  # Add OSM street map layer
  geom_point(data = shared_data, aes(x = longitude, y = latitude, color = case_num, size = case_num, 
                                     text = paste("Total Cases: ", case_num, " | Area Name: ", area_name))) +
  coord_sf(
    xlim = c(-79.6303856025883, -79.1180336515019),
    ylim = c(43.5822069220627, 43.8554571861712), expand = FALSE
  ) +
  theme_minimal() +
  labs(title = "Toronto Stolen Bike Case Numbers by Area", x = "Longitude", y = "Latitude", color = "Number of Cases", size = "Total Cases") +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_color_continuous(labels = scales::comma) +
  scale_size_continuous(range = c(3, 12)) # Adjust range values to control the size of points

>>>>

required_packages <- c("sf", "here", "dplyr", "readr", "purrr", "plotly", "ggplot2", "leaflet", "osmdata", 
                       "ggridges", "cowplot", "lubridate", "tidyverse", "crosstalk", "bookdown", "hrbrthemes", "gganimate")

for (p in required_packages) {
  if (!require(p, character.only = TRUE)) {
    install.packages(p, character.only = TRUE)
  }
}

library(sf)
library(here)
library(dplyr)
library(readr)
library(purrr)
library(plotly)
library(ggplot2)
library(leaflet)
library(osmdata)
library(cowplot)
library(ggridges)
library(bookdown)
library(lubridate)
library(tidyverse)
library(crosstalk)
library(gganimate)
library(hrbrthemes)

```

@@

@@

# Initialize plot with a line trace for the entire data
fig <- plot_ly(data, x = ~OCC_DOW, y = ~Number_of_Thefts, type = 'scatter', mode = 'lines+markers',
               line = list(color = '#05595B'), marker = list(color = '#603601', size = 10))

# Add points for each day with initial visibility settings
for (i in 1:nrow(data)) {
  fig <- fig %>%
    add_trace(
      x = data$OCC_DOW[i],
      y = data$Number_of_Thefts[i],
      type = 'scatter',
      mode = 'markers+text',
      text = paste("Thefts:", data$Number_of_Thefts[i]),
      textposition = "top center",
      marker = list(color = "#FFC107", size = 15),
      showlegend = FALSE,
      visible = ifelse(i == 1, TRUE, FALSE)
    )
}

@@

<div class="main-body">
  <p> Weekly bicycle theft patterns in Toronto is illustrated in \@ref(fig:interactive_weekly_plot), revealing that thefts are not evenly distributed throughout the week. The data shows a peak on Fridays, with over <strong style="color: #FFC107;">4,900</strong> reported thefts, followed by a decline on Saturday and Sunday. This trend may reflect how people use their bikes during the week versus the weekend. Fridays might represent a combination of commuter bike usage and weekend planning, where people may leave their bikes in high-traffic or less secure areas, making them more vulnerable to theft.</p>

@@

<center>
```{r mothly_plot, include = TRUE, warning = FALSE, message = FALSE, fig.width=9.5, fig.height=5}

@@

```{r}

```

@@

<center>
```{r animated_bar_plot, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=4.5, fig.height=2.3}

@@

<center>
```{r animated_line_plot, echo=FALSE, message=FALSE, warning=FALSE, results='asis', fig.width=5, fig.height=3}

@@

The year 2020, however, deviates from this regular trend due to the impact of the COVID-19 pandemic. As public transit options were considered risky, many Torontonians turned to bicycles as a safer and more flexible transportation option, resulting in a rise in bike ownership and driving up theft rates (@citeShapiro2021). Additionally, the pandemic stretched police resources, potentially limiting routine patrols and recovery efforts. The 2020 increase highlights how external factors, such as a public health crisis, can exacerbate bike theft trends and points to the importance of adaptable theft prevention strategies in response to changing urban transportation needs.

@@

# Load your dataset
map_data <- read_csv(here("data", "map_by_neighbourhoods_data.csv"), show_col_types = FALSE)

# Modify the data frame, categorizing case_num column
map_data <- map_data %>% 
  mutate(
    mag.level = cut(case_num, c(100, 500, 1000, 1500),
                    labels = c('> 100 & <=500', '>500 & <=1000', '>1000 & <=1500'))
  ) %>%
  split(.$mag.level)

# Initialize the leaflet map
l <- leaflet() %>%
  addProviderTiles(providers$Esri.OceanBasemap)

# Add the markers to the map for each case_num level
names(map_data) %>%
  walk(function(df) {
    l <<- l %>% 
      addMarkers(
        data = map_data[[df]], lng = ~longitude, lat = ~latitude,
        label = ~paste0("Hood: ", hood_name, " | Stolen Place: ", stolen_place, " | Cost: ", cost, " | Status: ", status),
        popup = ~paste0(
          "Hood: ", hood_name, "<br>",
          "Stolen Place: ", stolen_place, "<br>",
          "Cost: ", cost, "<br>",
          "Status: ", status, "<br>",
          "Case in area: ", case_num
        ),
        group = df,
        clusterOptions = markerClusterOptions()
      )
  })

# Add layer controls and a minimap
l <- l %>%
  addLayersControl(
    overlayGroups = names(map_data),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addMiniMap(tiles = providers$Esri.OceanBasemap, width = 120, height = 80)

# Display the Map
l

@@

<center>
```{r map_of_stolen_bike, echo=FALSE, warning=FALSE, message=FALSE}

@@

# Plot the map with bike theft locations data
bike_theft_plot <- ggplot() +
  geom_sf(data = streets, color = "gray", size = 0.3, alpha = 0.6) + # OSM street map layer
  geom_point(
    data = coordinates_data, aes(x = longitude, y = latitude),
    size = 0.5, color = "#FFC107"
  ) + # Bike theft points
  labs(
    title = "Map of Stolen Bike Cases in Toronto",
    x = "Longitude", y = "Latitude"
  ) +
  coord_sf(
    xlim = c(-79.6303856025883, -79.1180336515019),
    ylim = c(43.5822069220627, 43.8554571861712), expand = FALSE
  ) +
  theme_minimal() +
  theme(
    panel.background = element_blank(), # Remove panel background
    panel.grid.major = element_blank(), # Remove major grid lines
    panel.grid.minor = element_blank(), # Remove minor grid lines
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

bike_theft_plot

@@

@@

# Define more detailed breaks for color scale based on `case_num`
max_case_num <- max(map_data$case_num, na.rm = TRUE)
color_breaks <- seq(0, max_case_num, by = max_case_num / 4)

# Create a rounding function to simplify the labels
rounded_labels <- function(x) scales::comma(round(x, -6))

# Create ggplot object with the dataset and Toronto map as a background
ggplot_map <- ggplot() +
  geom_sf(data = streets, color = "lightgray", size = 0.3, alpha = 0.6) +  # Add OSM street map layer
  geom_point(data = shared_data, aes(x = longitude, y = latitude, color = case_num, size = case_num, 
                                     text = paste("Total Cases: ", case_num, " | Area Name: ", area_name))) +
  coord_sf(
    xlim = c(-79.6303856025883, -79.1180336515019),
    ylim = c(43.5822069220627, 43.8554571861712), expand = FALSE
  ) +
  theme_minimal() +
  labs(title = "Toronto Stolen Bike Case Numbers by Area", x = "Longitude", y = "Latitude", color = "Number of Cases", size = "Total Cases") +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_color_gradientn(colors = c("#603601", "#FFC107"), breaks = color_breaks, labels = rounded_labels) +
  scale_size_continuous(range = c(3, 12))
